CODECHECK: An Open Science initiative for the independent execution of computations underlying research articles during peer review improves reproducibility

### 2021-02-01 codecheck-paper.tex
manual page breaks

@@

\textbf{1. Codecheckers record but don't investigate or fix.} The
codechecker follows the instructions provided by the author in running
the code. If any steps are unclear, or if code does not run correctly,
the codechecker records this fact and then communicates this
information to the author. We believe that the job of the codechecker is
not to fix all these problems but simply to report them and await a
fix from the author.
The level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right, and too often this 
uncertainty leads researchers to give up and not document it at all.
The conversation with a codechecker fixes this problem.

\textbf{2. Communication between humans is key.} Some code may
work without any interaction (e.g.,~CODECHECK certificate \texttt{2020-013}
\cite{cert-2020-013}) but more often there are hidden
dependencies that need to be found or software installations that need to be adjusted for a particular system.
Allowing the (human) codechecker to communicate
directly and openly with each other should make this process as
constructive as possible. We feel that routing this conversation
(possibly anonymously) through a publisher would slow things down and
reduce the chances for community building as well as obscure some of the lessons
learned for both authors and codecheckers.

\textbf{3. Credit is given to codecheckers.} The value of performing a
CODECHECK is comparable to that of a scientific peer review, and
it may also require a similar amount of time. Therefore,
the codechecker's activity should be publicly recorded and mentioned in the
published paper.
The public record can be realised by publishing the certificate in a 
citable form (i.e.,~with a DOI, as conducted in the community CODECHECK
process), by listing codecheckers on the journal's website or, ideally, by
publishing the checks alongside peer review activities in public databases.

\textbf{4. Workflows must be auditable.}
This requires that the codechecker 
has at least enough material to validate the workflow outputs submitted by 
the authors. Stark~\cite{stark_before_2018} calls this ``preproducibility''
and the ICERM report \cite{stodden_setting_2013} defines the level
``Auditable Research'' similarly.
Every community may establish their own good practices or adapt generic
concepts and practical tools, such as publishing all building blocks
of science in a research compendium
\footnote{\url{https://research-compendium.science/}}.
A completed check means that code could be executed at least once without
critical errors or
warnings using the provided instructions, and, therefore, all code and data 
was given and could be investigated more deeply or extended in the future.
Ideally, this is a ``one~click'' step, but achieving this requires particular 
skills and a sufficient level of documentation for 
third parties. Furthermore, automation may lead to people gaming the system
or reliance on technology, which can often hide important details.
All such aspects can reduce the understandability of the material, so we estimate our approach to codechecking, done without automation and with open human communication, to be the best way to ensure long-term transparency and usefulness.
We acknowledge that others have argued in favour of bitwise reproducibility
because, in the long run, it can be automated\footnote{Konrad Hinsen on 
Twitter: \url{https://twitter.com/khinsen/status/1242842759733665799}},
but until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.}
By default, unless there are strong
reasons to the contrary (e.g.,~sensitive data on human subjects), all code and
data, both from author and codechecker, will be made freely available, 
i.e.,~under open licenses, before or when the certificate is published.
Note that openness is not required for the paper itself, as to accommodate 
journals in their transition to sustainable Open Access models.
The code and data publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed (a) with a recognition of
shortcomings in asserting reproducibility of submissions and (b)
with an understanding of the
complex situation that led to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives aiming to improve education
around reproducibility of methods and communication surrounding comprehensive 
computational workflows.
The long term goal is for a specific community or
journal's target audience to reach a state where the level of verification 
ensured by codecheckers becomes part of peer review and a regular
merit of all accepted papers.

@@

\textbf{Isn't CODECHECK what peer review should be doing already?}
On the surface, yes.
However, the general observation is that the
increasing number of researchers, submissions, and reviews to conduct
under the pressure to publish puts the scholarly publication system
under significant strain.
Miyakawa described how when he as an editor simply requested raw data from authors, this led to almost all of those papers in which data was requested to be withdrawn or rejected \cite{miyakawa_no_2020}.
Such basic checks to show that
sufficient data supports the results requires much effort from editors,
yet they only rely on the availability of data files and the content of the
paper.
These availability checks can be enhanced by having more complex CODECHECKs request the code and then execute it, because code is not inert but needs to be run in a specific environment.
This might fall within idealistic expectations of peer review,
but is far from common in reality.
Establishing a CODECHECK
process is largely an acknowledgement that peer reviewing practices
have been unable to adapt to the challenges of digitisation of data-based science.
The concept of a CODECHECK,
just as the concepts of reproducible research and Open Science,
may be transitional by nature. If the activities described here as
being part of a CODECHECK become the norm for scientific reviews and
publications, then the initiative will have succeeded. However, the
rapidly changing technology used in research leads us to believe that
a separate role of codechecker and the expertise it brings will be
beneficial in the long term.

\textbf{Aren't CODECHECK requirements not demanding enough?}
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approach even if such suggestions, e.g.,~research compendia\footnote{\url{https://research-compendium.science/}}, are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and the possibility that a workflow can be extended by third parties in the future, as long as they have access
to suitable resources and matching skill set.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors on how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow the bar to be raised as high as possible, e.g.,~with
executable research compendia \cite{nust_opening_2017}, ``Whole Tales'' \cite{brinckman_computing_2018},
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as key barriers to reproducibility
(e.g.,~\cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.
For example, ModelDB curation policies ensure that only one figure is manually reproduced \cite{mcdougal_reproducibility_2016}, but that has not limited the usefulness nor success of the platform.

>>>>

\newpage

@@

\textbf{1. Codecheckers record but don't investigate or fix.} \\
The codechecker follows the instructions provided by the author in running
the code. If any steps are unclear, or if code does not run correctly,
the codechecker records this fact and then communicates this
information to the author. We believe that the job of the codechecker is
not to fix all these problems but simply to report them and await a
fix from the author.
The level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right, and too often this 
uncertainty leads researchers to give up and not document it at all.
The conversation with a codechecker fixes this problem.

\textbf{2. Communication between humans is key.} \\ 
Some code may
work without any interaction (e.g., CODECHECK certificate \texttt{2020-013}
\cite{cert-2020-013}) but more often there are hidden
dependencies that need to be found or software installations that need to be adjusted for a particular system.
Allowing the (human) codechecker to communicate
directly and openly with each other should make this process as
constructive as possible. We feel that routing this conversation
(possibly anonymously) through a publisher would slow things down and
reduce the chances for community building as well as obscure some of the lessons
learned for both authors and codecheckers.

\textbf{3. Credit is given to codecheckers.} \\ 
The value of performing a
CODECHECK is comparable to that of a scientific peer review, and
it may also require a similar amount of time. Therefore,
the codechecker's activity should be publicly recorded and mentioned in the
published paper.
The public record can be realised by publishing the certificate in a 
citable form (i.e.,~with a DOI, as conducted in the community CODECHECK
process), by listing codecheckers on the journal's website or, ideally, by
publishing the checks alongside peer review activities in public databases.

\textbf{4. Workflows must be auditable.} \\ 
This requires that the codechecker 
has at least enough material to validate the workflow outputs submitted by 
the authors. Stark~\cite{stark_before_2018} calls this ``preproducibility''
and the ICERM report \cite{stodden_setting_2013} defines the level
``Auditable Research'' similarly.
Every community may establish their own good practices or adapt generic
concepts and practical tools, such as publishing all building blocks
of science in a research compendium
\footnote{\url{https://research-compendium.science/}}.
A completed check means that code could be executed at least once without
critical errors or
warnings using the provided instructions, and, therefore, all code and data 
was given and could be investigated more deeply or extended in the future.
Ideally, this is a ``one~click'' step, but achieving this requires particular 
skills and a sufficient level of documentation for 
third parties. Furthermore, automation may lead to people gaming the system
or reliance on technology, which can often hide important details.
All such aspects can reduce the understandability of the material, so we estimate our approach to codechecking, done without automation and with open human communication, to be the best way to ensure long-term transparency and usefulness.
We acknowledge that others have argued in favour of bitwise reproducibility
because, in the long run, it can be automated\footnote{Konrad Hinsen on 
Twitter: \url{https://twitter.com/khinsen/status/1242842759733665799}},
but until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.} \\ 
By default, unless there are strong
reasons to the contrary (e.g.,~sensitive data on human subjects), all code and
data, both from author and codechecker, will be made freely available, 
i.e.,~under open licenses, before or when the certificate is published.
Note that openness is not required for the paper itself, as to accommodate 
journals in their transition to sustainable Open Access models.
The code and data publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed (a) with a recognition of
shortcomings in asserting reproducibility of submissions and (b)
with an understanding of the
complex situation that led to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives aiming to improve education
around reproducibility of methods and communication surrounding comprehensive 
computational workflows.
The long term goal is for a specific community or
journal's target audience to reach a state where the level of verification 
ensured by codecheckers becomes part of peer review and a regular
merit of all accepted papers.

@@

\textbf{Isn't CODECHECK what peer review should be doing already?} \\
On the surface, yes.
However, the general observation is that the
increasing number of researchers, submissions, and reviews to conduct
under the pressure to publish puts the scholarly publication system
under significant strain.
Miyakawa described how when he as an editor simply requested raw data from authors, this led to almost all of those papers in which data was requested to be withdrawn or rejected \cite{miyakawa_no_2020}.
Such basic checks to show that
sufficient data supports the results requires much effort from editors,
yet they only rely on the availability of data files and the content of the
paper.
These availability checks can be enhanced by having more complex CODECHECKs request the code and then execute it, because code is not inert but needs to be run in a specific environment.
This might fall within idealistic expectations of peer review,
but is far from common in reality.
Establishing a CODECHECK
process is largely an acknowledgement that peer reviewing practices
have been unable to adapt to the challenges of digitisation of data-based science.
The concept of a CODECHECK,
just as the concepts of reproducible research and Open Science,
may be transitional by nature. If the activities described here as
being part of a CODECHECK become the norm for scientific reviews and
publications, then the initiative will have succeeded. However, the
rapidly changing technology used in research leads us to believe that
a separate role of codechecker and the expertise it brings will be
beneficial in the long term.

\newpage

\textbf{Should CODECHECK requirements not be more demanding?} \\
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approach even if such suggestions, e.g.,~research compendia\footnote{\url{https://research-compendium.science/}}, are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and the possibility that a workflow can be extended by third parties in the future, as long as they have access
to suitable resources and matching skill set.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors on how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow the bar to be raised as high as possible, e.g.,~with
executable research compendia \cite{nust_opening_2017}, ``Whole Tales'' \cite{brinckman_computing_2018},
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as key barriers to reproducibility
(e.g.,~\cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.
For example, ModelDB curation policies ensure that only one figure is manually reproduced \cite{mcdougal_reproducibility_2016}, but that has not limited the usefulness nor success of the platform.
